---
title: "Interactions from function"
author: "Stefano Allesina"
date: '2022-07-09'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting

We have a pool of $n$ species, with dynamics governed by GLV:

$$
\frac{d x_i}{dt} = x_i r_i (1 - \sum_j B_{ij} x_j)
$$

We perform experiments in which a subset $k$ of species are grown together. If the species can stably coexist (either at equilibrium, or in a limit cycle or chaotic attractor), their time-averaged abundance will be:

$$
x^{(k)} = \left( B^{(k)}\right)^{-1} 1
$$
where $x^{(k)}$ is a vector containing the densities of the species in $k$, and $B^{(k)}$ is the sub-matrix of $B$ in which we have retained only the rows/cols corresponding to the species in $k$.

We perform several experiments, each time varying the initial composition. For each experiment, we measure a function $f^{(k)}$ (for example, total biomass, respiration, production of a certain metabolite). We make a strong assumption:

$$
f^{(k)} = \sum_{j \in k} \alpha_j x_{j}^{(k)}
$$

The goal of the project is to predict $f^{(k)}$ for novel communities, when we have learned the relevant parameters from a set of experiments. 

## Change of variables

We have:

$$
f^{(k)} = \sum_{i,j} \alpha_i (B^{(k)})^{-1}_{ij}  = (\alpha^{(k)})^T (B^{(k)})^{-1} 1
$$

Call $B D(\frac{1}{\alpha}) = Q$. Then 

$$
f^{(k)} = 1^T (B^{(k)}D(\frac{1}{\alpha^{(k)}}))^{-1} 1 = 1^T D(\alpha^{(k)}) (B^{(k)})^{-1}1 = 1^T (Q^{(k)})^{-1}1
$$

Because we always have that $f^{(k)}$ is a quadratic form, only the symmetric part of $(Q^{(k)})^{-1}$ will matter. As such, the simplest case to study is that in which $Q$ is symmetric.

### Learning the parameters for symmetric Q

Suppose $n=3$, and that we have performed experiments spanning all the $2^n -1$ combinations. We want to learn

$$
Q = \begin{pmatrix}
q_{11} & q_{12} & q_{13}\\
q_{12} & q_{22} & q_{23}\\
q_{13} & q_{23} & q_{33}
\end{pmatrix}
$$


Analyzing the three experiments with a single species we get:

$$
\begin{aligned}
(Q^{(1)})^{-1} & = \frac{1}{q_{11}} = f^{(1)}\\
(Q^{(2)})^{-1} & = \frac{1}{q_{22}} = f^{(2)}\\
(Q^{(3)})^{-1} & = \frac{1}{q_{33}} = f^{(3)}\\
\end{aligned}
$$

Solving:

$$
\begin{aligned}
q_{11} = \frac{1}{f^{(1)}}\\
q_{22} = \frac{1}{f^{(2)}}\\
q_{33} = \frac{1}{f^{(3)}}\\
\end{aligned}
$$

Now take the two-species experiments:

$$
f^{(i,j)} = 1^T \begin{pmatrix}
q_{ii} & q_{ij}\\
q_{ij} & q_{jj}
\end{pmatrix}^{-1} 1 = 1^T \begin{pmatrix}
\frac{1}{f^{(i)}} & q_{ij}\\
q_{ij} & \frac{1}{f^{(j)}}
\end{pmatrix}^{-1} 1 =  \frac{-2 f^{(i)} f^{(j)} q_{ij}+ f^{(i)}+ f^{(j)}}{1-f^{(i)} f^{(j)} q_{ij}^2}
$$

We can solve for $q_{ij}$, finding two solutions:

$$
q_{ij} = \frac{1 \pm \sqrt{\frac{(f^{(i,j)}-f^{(i)}) (f^{(i,j)}-f^{(j)})}{f^{(i)} f^{(j)}}}}{f^{(i,j)}}
$$

Given that we have three pairs $i,j$, we end up with 8 possible solutions. Problem: how do we choose the right solution? If we have a solution for all the coefficients in $Q$, we can attempt predicting the function $f^{(1,2,3)}$.

### Example code

```{r}
set.seed(12)
n <- 3
# make Q symmetric
B <- matrix(-rnorm(n * n), n, n)
alpha <- runif(n)
Q <- B %*% diag(1 / alpha)
Q <- Q + t(Q)
B <- Q %*% diag(alpha)

# compute function for each sub-community, using either B, alpha or Q
results <- matrix(0, 0, n + 2)
for (i in 1:(2^n - 1)){
  presence <- as.numeric(intToBits(i)[1:n])
  Bk <- B[presence > 0, presence > 0, drop = FALSE]
  alphak <- alpha[presence > 0]
  Qk <- Q[presence > 0, presence > 0, drop = FALSE]
  results <- rbind(results, c(presence, alphak %*% rowSums(solve(Bk)), sum(solve(Qk))))
}
# note that the results are the same
print(results)
# solve system  of equations
f <- as.complex(results[,n+1]) # the part under sqrt can be negative...
f1 <- f[1]
f2 <- f[2]
f3 <- f[4] # this is spp 3 by itself
f12 <- f[3] # this is spp 1 and 2
f13 <- f[5] 
f23 <- f[6] 
k11 <- 1/f1
k22 <- 1/f2
k33 <- 1/f3
k12 <- (1 + sqrt((f12 - f1) * (f12 - f2) / (f1 * f2))) / f12 # choose +
k13 <- (1 - sqrt((f13 - f1) * (f13 - f3) / (f1 * f3))) / f13 # choose -
k23 <- (1 + sqrt((f23 - f2) * (f23 - f3) / (f2 * f3))) / f23 # choose +
K <- matrix(c(k11, k12, k13,
              k12, k22, k23,
              k13, k23, k33), 3, 3, byrow = TRUE)
# we have recovered all the coefficients
print(Q)
print(Re(K))
# we can therefore predict the function for all spp together
sum(solve(Re(K)))
```

## Generic Q

```{r}
get_results <- function(Q){
  n <- nrow(Q)
  results <- matrix(0, 0, n + 1)
  for (i in 1:(2^n - 1)){
    presence <- as.numeric(intToBits(i)[1:n])
    Qk <- Q[presence > 0, presence > 0, drop = FALSE]
    results <- rbind(results, c(presence, sum(solve(Qk))))
  }
  return(results)
}
```


Can we learn $Q$ when it is not symmetric?

```{r}
set.seed(13)
n <- 5
# make Q non symmetric
Q <- abs(matrix(rnorm(n * n), n, n))
diag(Q) <- diag(Q) + 1 # strong diagonal
results <- get_results(Q)
design_matrix <- results[,1:n]
f <- results[,n+1]

get_ssq <- function(pars, design_matrix, f, returnf = FALSE){
  n <- sqrt(length(pars))
  Q <- matrix(pars, n, n)
  fQ <- f
  for (i in 1:length(f)){
    presence <- design_matrix[i,]
    Qk <- Q[presence > 0, presence > 0, drop = FALSE]
    fQ[i] <- sum(solve(Qk))
  }
  if (returnf) return(fQ)
  return(sum((f - fQ)^2))
}

hc <- function(bestpars, design_matrix, f, nsteps = 500, startpert = 0.25){
  bestres <- get_ssq(bestpars, design_matrix, f)
  dev <- startpert
  for (j in 1:10){
    for (i in 1:nsteps){
      tmppar <- bestpars * (1 + rnorm(length(bestpars)) * dev)
      tmpres <- get_ssq(tmppar, design_matrix, f)
      if (tmpres < bestres){
        bestpars <- tmppar
        bestres <- tmpres
        #print(tmpres)
      }
    }
    dev <- dev / 2
  }
  return(bestpars)
}

tmp <- list(par = as.vector(diag(rep(1,n)))) # start with identity matrix
for (iter in 1:10){
  print(iter)
  tmp$par <- hc(tmp$par, design_matrix, f)
  tmp$value <- get_ssq(tmp$par, design_matrix, f)
  
  tmp <- optim(par = tmp$par, fn = get_ssq, design_matrix = design_matrix, f = f, 
             method = "Nelder-Mead", control = list(maxit = 5000, trace = FALSE))
  print(tmp$value)
  print(tmp$value)
  tmp <- optim(par = tmp$par, fn = get_ssq, design_matrix = design_matrix, f = f, 
             method = "BFGS", control = list(maxit = 5000, trace = FALSE))
  print(tmp$value)
}
fQ <- get_ssq(tmp$par, design_matrix, f, TRUE)
plot(f, fQ)
abline(c(0,1))
K <- matrix(as.vector(tmp$par), n, n)
plot(as.vector(Re(K)), as.vector(Q))
abline(c(0,1))
```


